---
title: "Crinier_dataset_Preprocessing"
author: "Dillon Corvino"
date: "13/06/2019"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# High-Dimensional Single-cell analysis identifies organ-specific signatures and conserved NK cell subsets in Humans and Mice
# Crinier et al., 2018, Immunity 

# Organs: Blood and Spleen 
# Species: Human and Mouse


# 6 human samples
# 6 mouse samples

# samples 1-3 = Spleen
# samples 4-6 = Blood


```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Load packages
source("R/Load_packages.R", local = knitr::knit_global())

# Create output directories & load custom functions & colour scheme
source("R/Setup.R", local = knitr::knit_global())

```

### Read and format Human samples
```{r reading_data_human}

# List files 
data.list <- list.files("Data/", pattern = ".csv")

# Identify human files
human.logic <- grepl("sample.h", data.list)

human.files <- data.list[human.logic]

#################
# Human dataset 
#################

# Create name variable
file.names <- gsub(".*_", "", human.files)
file.names <- gsub(".csv", "", file.names)

# Read in human samples
for (i in seq_along(human.files)){
  assign(file.names[i], read.csv(paste0("Data/", human.files[i])))
}


# Merge data by organ of origin
spleen.raw <- 
  full_join(sample1h, sample2h,  by = "X") %>%
  full_join(sample3h,  by = "X") %>%
  arrange(X, desc = T)
  
blood.raw <- 
  full_join(sample4h, sample5h,  by = "X") %>%
  full_join(sample6h,  by = "X") %>%
  arrange(X, desc = T)


# Checking all genes were merged 
spleen.geneIDs <- unique(c(as.character(sample1h$X), 
                          as.character(sample2h$X), 
                          as.character(sample3h$X)))
length(spleen.geneIDs) == nrow(spleen.raw)

blood.geneIDs <- unique(c(as.character(sample4h$X), 
                          as.character(sample5h$X), 
                          as.character(sample6h$X)))

length(blood.geneIDs) == nrow(blood.raw)

# overview stats
dim(spleen.raw)
dim(blood.raw)

#####################
# Format dataframe
#####################

# Put gene ID as rownames
rownames(spleen.raw) <- as.character(spleen.raw$X)
rownames(blood.raw) <- as.character(blood.raw$X)

# remove former gene ID column
index.val <- grep("^X$", colnames(spleen.raw))
colnames(spleen.raw)[index.val]
spleen.raw <- spleen.raw[,-index.val]


index.val <- grep("^X$", colnames(blood.raw))
colnames(blood.raw)[index.val]
blood.raw <- blood.raw[,-index.val]


# Annotate the cell Ids so can be defined later
colnames(spleen.raw) <- paste0(colnames(spleen.raw), "_Sp")
colnames(blood.raw) <- paste0(colnames(blood.raw), "_Bl")


# Merge all data together
human.data <- merge(spleen.raw, blood.raw, by = "row.names", all = TRUE)

# Annotate
rownames(human.data) <- human.data$Row.names

index.val <- grep("^Row.names$", colnames(human.data))
colnames(human.data)[index.val]
human.data <- human.data[,-index.val]


# Correct number of columns 
total.geneIDs <- unique(c(spleen.geneIDs, 
                          blood.geneIDs))

length(total.geneIDs) == nrow(human.data)

# Remove NA values
na.values <- is.na(human.data)
sum(na.values) # 19,966,040

# Assigning NA values as 0
human.data[na.values] <- 0

# Checking NA values are removed
sum(is.na(human.data)) 

# Checking for any merging conflicts
# Found that some duplicate annoation for 7 cells in the blood
colnames(human.data)[grep(".x", colnames(human.data))]
colnames(human.data)[grep(".y", colnames(human.data))]

# Remove these duplicated columns
x.index <- grep(".x", colnames(human.data))
y.index <- grep(".y", colnames(human.data))
human.data <- human.data[, -c(x.index, y.index)]

dim(human.data) # 15,657 genes and 8,480 cells 


# Write data to file for easy use later
write.csv(human.data, "output/tables/Human_Sp_and_Bl_samples_raw.csv")

# remove large datafiles
rm(sample1h, sample2h,
   sample3h, sample4h,
   sample5h, sample6h, 
   spleen.raw, human.data,
   blood.raw)

```

### Read and format Mouse samples
```{r reading_data_mouse}

# List files 
data.list <- list.files("Data/", pattern = ".csv")

# Identify mouse files
mouse.logic <- grepl("sample.m", data.list)

mouse.files <- data.list[mouse.logic]

#################
# Mouse dataset 
#################

# Create name variable
file.names <- gsub(".*_", "", mouse.files)
file.names <- gsub(".csv", "", file.names)

# Read in human samples
for (i in seq_along(mouse.files)){
  assign(file.names[i], read.csv(paste0("Data/", mouse.files[i])))
}


# Merge data by organ of origin
spleen.raw <- 
  full_join(sample1m, sample2m,  by = "X") %>%
  full_join(sample3m,  by = "X") %>%
  arrange(X, desc = T)
  
blood.raw <- 
  full_join(sample4m, sample5m,  by = "X") %>%
  full_join(sample6m,  by = "X") %>%
  arrange(X, desc = T)

# Checking all genes were merged 
spleen.geneIDs <- unique(c(as.character(sample1m$X), 
                          as.character(sample2m$X), 
                          as.character(sample3m$X)))

length(spleen.geneIDs) == nrow(spleen.raw)

blood.geneIDs <- unique(c(as.character(sample4m$X), 
                          as.character(sample5m$X), 
                          as.character(sample6m$X)))

length(blood.geneIDs) == nrow(blood.raw)

# overview stats
dim(spleen.raw)
dim(blood.raw)

#####################
# Format dataframe
#####################

# Put gene ID as rownames
rownames(spleen.raw) <- as.character(spleen.raw$X)
rownames(blood.raw) <- as.character(blood.raw$X)

# remove former gene ID column
index.val <- grep("^X$", colnames(spleen.raw))
colnames(spleen.raw)[index.val]
spleen.raw <- spleen.raw[,-index.val]


index.val <- grep("^X$", colnames(blood.raw))
colnames(blood.raw)[index.val]
blood.raw <- blood.raw[,-index.val]


# Annotate the cell Ids so can be defined later
colnames(spleen.raw) <- paste0(colnames(spleen.raw), "_Sp")
colnames(blood.raw) <- paste0(colnames(blood.raw), "_Bl")


# Merge all data together
mouse.data <- merge(spleen.raw, blood.raw, 
                    by = "row.names", 
                    all = TRUE)

# Annotate
rownames(mouse.data) <- mouse.data$Row.names

index.val <- grep("^Row.names$", colnames(mouse.data))
colnames(mouse.data)[index.val]
mouse.data <- mouse.data[,-index.val]


# Correct number of columns 
total.geneIDs <- unique(c(spleen.geneIDs, 
                          blood.geneIDs))

length(total.geneIDs) == nrow(mouse.data)

# Remove NA values
na.values <- is.na(mouse.data)
sum(na.values) # 10,686,471

# Assigning NA values as 0
mouse.data[na.values] <- 0

# Checking NA values are removed
sum(is.na(mouse.data)) 

# Checking for any merging conflicts
# found some duplicate cell annotations
colnames(mouse.data)[grep(".x", colnames(mouse.data))]
colnames(mouse.data)[grep(".y", colnames(mouse.data))]

# Remove these duplicated columns
x.index <- grep(".x", colnames(mouse.data))
y.index <- grep(".y", colnames(mouse.data))
mouse.data <- mouse.data[, -c(x.index, y.index)]

dim(mouse.data) # 12,150 genes and 8,203 cells 


# Write data to file for easy use later
write.csv(mouse.data, "output/tables/Mouse_Sp_and_Bl_samples_raw.csv")

# remove large datafiles
rm(sample1m, sample2m,
   sample3m, sample4m,
   sample5m, sample6m, 
   spleen.raw, mouse.data,
   blood.raw)

```

# Uses input from previous code chunk
```{r create_Seurat_object}

human.data <- read.csv("output/tables/Human_Sp_and_Bl_samples_raw.csv")
dim(human.data)
human.data[1:10, 1:10]

rownames(human.data) <- human.data$X
human.data <- human.data[, -1]


#spleen.logic <- grepl("_Sp", colnames(human.data))
#blood.logic <- grepl("_Bl", colnames(human.data))

#spleen.data <- human.data[, spleen.logic]
#blood.data <- human.data[, blood.logic]




temp.data <- read.csv("Data/Metadata/Mouse2Blood_metadata.csv")
head(temp.data)
tail(temp.data)



data.list <- list.files("Data/Metadata", pattern = ".csv")


#################
# Mouse dataset 
#################

# Create name variable
file.names <- gsub("_.*", "", data.list)
file.names <- gsub(".csv", "", file.names)

# Read in human samples
for (i in seq_along(data.list)){
  assign(file.names[i], read.csv(paste0("Data/Metadata/", data.list[i])))
}


# Initialize the Seurat object with the raw (non-normalized data).  
# Keep all genes expressed in >= 3 cells. 
# Keep all cells with at least 200 detected genes

human.seurat <- CreateSeuratObject(counts = human.data, 
                                    min.cells = 3, 
                                    min.genes = 200,
                                    is.expr = 0, # for UMI datasets this should be 0, and is default
                                    project = "Human_NK",
                                    display.progress = TRUE)


human.seurat

saveRDS(human.seurat, "Exported_RDS_files/Human_raw_Seurat_Object.RData")

```


## Analysis {.tabset}
### QC & Normalisation
```{r QC_and_normalisation}

################################################
#  Get percentage Mitochondria gene expression
################################################



species.ori <- "human"

if (species.ori == "mouse") {
  pat <- "^mt-"
}else if (species.ori == "human"){
  pat <- "^MT-"
}
  

human.seurat[["percent.mito"]] <- PercentageFeatureSet(human.seurat, pattern = paste0(pat))


VlnPlot(human.seurat, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), 
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nFeature_nCount_RNA_percentMito.pdf")
dev.off()


VlnPlot(human.seurat, 
              features = "percent.mito", 
              pt.size = 0.1, 
              y.max = 20) 

dev.copy(pdf, "output/QC/QC_percentMito.pdf")
dev.off()



head(human.seurat@meta.data)
str(human.seurat@meta.data)

#  Filter cells
# Remove cells with < 200 genes, and remove cells with > 10% mitochondrial genes

human.seurat.filt <- subset(human.seurat, 
                             subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mito < 10)





# Vis after filtering
VlnPlot(human.seurat,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), 
        pt.size = 0.1)

VlnPlot(human.seurat.filt, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), 
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nFeature_nCount_RNA_percentMito_Post_filtering.pdf")
dev.off()


# Mitopercent
p1 <- VlnPlot(human.seurat, 
              features = "percent.mito", 
              pt.size = 0.1, 
              y.max = 20) + NoLegend() + 
  ggtitle("Before filtering")

p2 <- VlnPlot(human.seurat.filt, 
              features = "percent.mito", 
              pt.size = 0.1, 
              y.max = 20) + NoLegend() + 
  ggtitle("After filtering")

p1 | p2

dev.copy(pdf, "output/QC/QC_percentMito_filtering.pdf")
dev.off()

# Stats
human.seurat # 12,236 genes across 8,480 cells
human.seurat.filt # 12,236 genes across 83,51 cells

# Therefore
8480 - 8351
# 3129 cells were deleted


####################################
# Annotating data by organ origin
####################################

#  Add organ label to metadata
organ <- rownames(human.seurat.filt@meta.data)
organ <- gsub(".*_","", organ)
organ <- gsub("Sp","Spleen", organ)
organ <- gsub("Bl","Blood", organ)
human.seurat.filt@meta.data$organ <- organ



# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("nFeature_RNA"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nFeature_postfiltering_byorgan.pdf")
dev.off()


# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("nCount_RNA"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nCount_postfiltering_byorgan.pdf")
dev.off()


# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("percent.mito"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_percentMito_postfiltering_byorgan.pdf")
dev.off()


################################################
# Normalise, scale & find variable genes 
################################################

# Normalise RNA data by logNormalize
human.seurat.filt <- NormalizeData(human.seurat.filt,
                                    normalization.method = "LogNormalize",
                                    scale.factor = 10000,
                                    verbose = TRUE)


# Find variable genes
human.seurat.filt <- FindVariableFeatures(human.seurat.filt,
                                           nfeatures = 1000,
                                           verbose = TRUE)

human.seurat.filt <- ScaleData(human.seurat.filt,
                                vars.to.regress = c("nCount_RNA", "percent.mito"),
                                model.use = "linear",
                                do.scale = TRUE,
                                do.center = TRUE,
                                verbose = TRUE)


###############################
# Remove unneeded variables
###############################

rm(human.seurat,
   human.data)

```



## Unbiased Clustering {.tabset}

### Clustering
```{r Initial_clustering}

# Create output dir
if(!dir.exists("output/figures/PCA")){
  dir.create("output/figures/PCA", 
             recursive = T)}



# Set seed for reproducibility
set.seed(42)

###########
# Run PCA
###########
human.seurat.filt <- RunPCA(human.seurat.filt, 
                             features = human.seurat.filt@assays$RNA@var.features,
                             npcs = 50,
                             verbose = TRUE, 
                             rev.pca = TRUE,
                             seed.use = 42)


#########################
# Plot PCA projection
#########################

PCAPlot(human.seurat.filt, 
        dims = c(1, 2),
        pt.size = 2,
        group.by = "organ")

dev.copy(pdf, "output/figures/PCA/PCA_by_organ_Dim1_2.pdf")
dev.off()

PCAPlot(human.seurat.filt, 
        dims = c(2, 3),
        pt.size = 2,
        group.by = "organ")

dev.copy(pdf, "output/figures/PCA/PCA_by_organ_Dim2_3.pdf")
dev.off()


#################################################################
# Determine statistically significant principal components
#################################################################

# NOTE: This process can take a long time for big datasets, comment out for
# expediency.  More approximate techniques such as those implemented in
# PCElbowPlot() can be used to reduce computation time

if(long.compute){
  
  human.seurat.filt <- JackStraw(human.seurat.filt,
                                  dims = 50)
  human.seurat.filt <- ScoreJackStraw(human.seurat.filt,
                                       dims = 1:50)
  
  JackStrawPlot(human.seurat.filt, dims = 1:30)
  
  dev.copy(pdf, "output/QC/JackStrawPlot.pdf")
  dev.off()
  
}

# PC 24 = last sig dim


# Use quick approximation of significant PCs
ElbowPlot(human.seurat.filt, ndims = 20)
dev.copy(pdf, "output/QC/PC_elbowplot.pdf")
dev.off()


##########################
# Find clusters using PCA
##########################

human.seurat.filt <- FindNeighbors(human.seurat.filt,
                                    dims = 1:20)

human.seurat.filt <- FindClusters(human.seurat.filt,
                                   resolution = 0.1)


# Resolution of 0.1 gives 5 communities

```


### UMAP and tSNE dim Reduction
```{r Dim_reduction}

# Create output dir
if(!dir.exists("output/figures/tSNE")){
  dir.create("output/figures/tSNE", 
             recursive = T)}

if(!dir.exists("output/figures/UMAP")){
  dir.create("output/figures/UMAP", 
             recursive = T)}


# Set assay 
DefaultAssay(human.seurat.filt) <- "RNA"

# Set Idents
Idents(human.seurat.filt) <- human.seurat.filt@meta.data$organ

##########################
# tSNE dim reduction
##########################

# tSNE perplex looks nice at 30

human.seurat.filt <- RunTSNE(human.seurat.filt,
                              reduction = "pca",
                              dims = 1:20,
                              perplexity = 30,
                              method = "FIt-SNE",
                              seed.use = 42,
                              dim.embed = 2)

TSNEPlot(human.seurat.filt,
         label = TRUE, 
         group.by = "organ")

dev.copy(pdf, "output/figures/tSNE/tSNE_group_by_organ.pdf")
dev.off()



TSNEPlot(human.seurat.filt,
         label = TRUE, 
         group.by = "seurat_clusters")

dev.copy(pdf, "output/figures/tSNE/tSNE_group_by_seurat_clusters.pdf")
dev.off()


##########################
# UMAP dim reduction
##########################
# UMAP min dis looks nice at = 0.5
# and nval = 45

human.seurat.filt <- RunUMAP(human.seurat.filt,
                              reduction = "pca",
                              dims = 1:20,
                              umap.method = "uwot", 
                              n.neighbors = 45, # 5 to 50 // 45
                              min.dist = 0.5, # Sensible values are in the range 0.001 to 0.5 // 0.5
                              seed.use = 42)

# Larger values = global structure
# small value = local structure

UMAPPlot(human.seurat.filt,
         label = T,
         group.by = "organ")

dev.copy(pdf, "output/figures/UMAP/UMAP_group_by_organ.pdf")
dev.off()


UMAPPlot(human.seurat.filt,
         label = TRUE, 
         group.by = "seurat_clusters")

dev.copy(pdf, "output/figures/UMAP/UMAP_group_by_seurat_clusters.pdf")
dev.off()


```


### Export seurat object
```{r export_seurat_RDS}

saveRDS(human.seurat.filt, file = "Exported_RDS_files/human_seurat_filt.rds")

```



## Imputation {.tabset}

### Imputation calculation
```{r seurat_imputation}

# Impute gene expression 
human.seurat.filt <- RunALRA(human.seurat.filt, 
                              genes.use = rownames(human.seurat.filt))

#The matrix went from 5.61% nonzero to 53.96% nonzero

# set assay 
DefaultAssay(human.seurat.filt) <- "RNA"

```

### Export seurat object - imputed
```{r export_imputed_RDS}

saveRDS(human.seurat.filt, file = "Exported_RDS_files/human_seurat_filt.rds")

```

