---
title: "Crinier_dataset_Human_Preprocessing"
author: "Dillon Corvino"
date: "13/06/2019"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# High-Dimensional Single-cell analysis identifies organ-specific signatures and conserved NK cell subsets in Humans and Mice
# Crinier et al., 2018, Immunity 

# Organs: Blood and Spleen 
# Species: Human and Mouse


# 6 human samples
# 6 mouse samples

# samples 1-3 = Spleen
# samples 4-6 = Blood


```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Load packages
source("R/Load_packages.R", local = knitr::knit_global())

# Create output directories & load custom functions & colour scheme
source("R/Setup.R", local = knitr::knit_global())

long.compute <- FALSE
quick.load <- TRUE

```


### Read and format Human samples
```{r reading_data_human}

# List files 
data.list <- list.files("Data/Human/Human_samples", pattern = ".csv")


# Create name variable
file.names <- gsub(".*_", "", data.list)
file.names <- gsub(".csv", "", file.names)


# Read in human samples
for (i in seq_along(data.list)){
  assign(file.names[i], read.csv(paste0("Data/Human/Human_samples/", data.list[i]), row.names = 1))
}




# Annotate the cell ids so can be identified later
colnames(sample1h) <- paste0(colnames(sample1h), "_Sp1")
colnames(sample2h) <- paste0(colnames(sample2h), "_Sp2")
colnames(sample3h) <- paste0(colnames(sample3h), "_Sp3")

colnames(sample4h) <- paste0(colnames(sample4h), "_Bl1")
colnames(sample5h) <- paste0(colnames(sample5h), "_Bl2")
colnames(sample6h) <- paste0(colnames(sample6h), "_Bl3")


########################################
# Merge data by organ of origin
########################################

# Dplyr doesnt allow merge by rowname. so have to make it a variable and then return the variable

# Spleen dataset
sample1h$GeneID <- row.names(sample1h)
sample2h$GeneID <- row.names(sample2h)
sample3h$GeneID <- row.names(sample3h)


spleen.raw <- 
  full_join(sample1h, sample2h,  by = "GeneID") %>%
  full_join(sample3h,  by = "GeneID") %>%
  arrange(desc = TRUE) %>%
  column_to_rownames(var = "GeneID")

# Blood dataset
sample4h$GeneID <- row.names(sample4h)
sample5h$GeneID <- row.names(sample5h)
sample6h$GeneID <- row.names(sample6h)

blood.raw <- 
  full_join(sample4h, sample5h,  by = "GeneID") %>%
  full_join(sample6h,  by = "GeneID") %>%
  arrange(desc = TRUE) %>%
  column_to_rownames(var = "GeneID")



# Checking all genes were merged 
spleen.geneIDs <- unique(c(as.character(sample1h$GeneID), 
                           as.character(sample2h$GeneID), 
                           as.character(sample3h$GeneID)))

length(spleen.geneIDs) == nrow(spleen.raw)

blood.geneIDs <- unique(c(as.character(sample4h$GeneID), 
                          as.character(sample5h$GeneID), 
                          as.character(sample6h$GeneID)))

length(blood.geneIDs) == nrow(blood.raw)

# overview stats
dim(spleen.raw)
dim(blood.raw)



#####################
# Format dataframe
#####################

# Merge all data together
human.data <- merge(spleen.raw, blood.raw, 
                    by = "row.names", 
                    all = TRUE)

# Annotate
rownames(human.data) <- human.data$Row.names

index.val <- grep("^Row.names$", colnames(human.data))
colnames(human.data)[index.val]
human.data <- human.data[,-index.val]


# Correct number of columns 
total.geneIDs <- unique(c(spleen.geneIDs, 
                          blood.geneIDs))

length(total.geneIDs) == nrow(human.data)



# When a gene was detected in spleen or blood but not the other there are now NA values for this entry in the other dataset cells
# Remove NA values
na.values <- is.na(human.data)
sum(na.values) # 19,966,040

# Assigning NA values as 0
human.data[na.values] <- 0

# Checking NA values are removed
sum(is.na(human.data)) 

dim(human.data) # 15,657 genes and 8,494 cells 

# Write data to file for easy use later
write.csv(human.data, "output/tables/Human_Sp_and_Bl_samples_raw.csv")
saveRDS(human.data, "Exported_RDS_files/Human_Sp_and_Bl_samples_raw.RDS")
saveRDS(spleen.raw, "Exported_RDS_files/Human_Sp_samples_raw.RDS")
saveRDS(blood.raw, "Exported_RDS_files/Human_Bl_samples_raw.RDS")




# remove large datafiles
rm(sample1h, sample2h,
   sample3h, sample4h,
   sample5h, sample6h, 
   spleen.raw, human.data,
   blood.raw)

```


# Uses input from previous code chunk
```{r create_Seurat_object}

human.data <- readRDS("Exported_RDS_files/Human_Sp_and_Bl_samples_raw.RDS")



# Initialize the Seurat object with the raw (non-normalized data).  
# Keep all genes expressed in >= 3 cells. 
# Keep all cells with at least 200 detected genes

human.seurat <- CreateSeuratObject(counts = human.data,
                                   min.cells = 3, 
                                   min.genes = 200,
                                   is.expr = 0, # for UMI datasets this should be 0, and is default
                                   project = "Human_NK",
                                   display.progress = TRUE)


# Create metadata
donor.ID <- gsub(".*_", "", rownames(human.seurat@meta.data))
organ <- gsub("[[:digit:]]", "", donor.ID)
organ <- gsub("Sp", "Spleen", organ)
organ <- gsub("Bl", "Blood", organ)
unique(organ)


human.seurat@meta.data$donor <- donor.ID
human.seurat@meta.data$organ <- organ




saveRDS(human.seurat, "Exported_RDS_files/Human_raw_Seurat_Object.RData")


# Remove unneeded data
rm(human.data)

```


## Analysis {.tabset}
### QC & Normalisation
```{r QC_and_normalisation}


################################################
#  Get percentage Mitochondria gene expression
################################################

species.ori <- "human"

if (species.ori == "mouse") {
  pat <- "^mt-"
}else if (species.ori == "human"){
  pat <- "^MT-"
}


human.seurat[["percent.mito.recalc"]] <- PercentageFeatureSet(human.seurat, pattern = paste0(pat))


VlnPlot(human.seurat, 
        features = c("nFeature_RNA", "nCount_RNA"),
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nFeature_nCount_RNA_percentMito.pdf")
dev.off()


VlnPlot(human.seurat, 
        features = c("percent.mito.recalc"), 
        pt.size = 0.1)


dev.copy(pdf, "output/QC/QC_percentMito.pdf")
dev.off()


#  Filter cells
# Remove cells with < 200 genes, and remove cells with > 10% mitochondrial genes
human.seurat.filt <- subset(human.seurat, 
                             subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mito.recalc < 10)



# Vis after filtering

p1 <- VlnPlot(human.seurat, 
              features = "percent.mito.recalc", 
              pt.size = 0.1, 
              y.max = 20) + NoLegend() + 
  ggtitle("Before filtering")

p2 <- VlnPlot(human.seurat.filt, 
              features = "percent.mito.recalc", 
              pt.size = 0.1, 
              y.max = 20) + NoLegend() + 
  ggtitle("After filtering")

p1 | p2

dev.copy(pdf, "output/QC/QC_percentMito_filtering.pdf")
dev.off()



p1 <- VlnPlot(human.seurat, 
              features = "nFeature_RNA", 
              pt.size = 0.1, 
              y.max = 2500) + NoLegend() + 
  ggtitle("Before filtering")

p2 <- VlnPlot(human.seurat.filt, 
              features = "nFeature_RNA", 
              pt.size = 0.1, 
              y.max = 2500) + NoLegend() + 
  ggtitle("After filtering")

p1 | p2

dev.copy(pdf, "output/QC/QC_nFeature_RNA_filtering.pdf")
dev.off()



# Stats
human.seurat # 15,657 genes across 8,494 cells
human.seurat.filt # 15,657 genes across 8,364 cells

# Therefore
8494 - 8364
# 130 cells were deleted




# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("nFeature_RNA"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nFeature_postfiltering_byorgan.pdf")
dev.off()


# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("nCount_RNA"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_nCount_postfiltering_byorgan.pdf")
dev.off()


# QC plots subset by organ assignment
VlnPlot(human.seurat.filt, 
        features = c("percent.mito.recalc"), 
        group.by = "organ",
        pt.size = 0.1)

dev.copy(pdf, "output/QC/QC_percentMito_postfiltering_byorgan.pdf")
dev.off()



# Remove unneeded variables
rm(human.seurat)

```

### Filtering dataset and subsetting dataset 
```{r filtering_and_subsetting_dataset}

##################
# Filter dataset 
##################


# Split dataset by organ
Idents(human.seurat.filt) <- human.seurat.filt@meta.data$organ


spleen.seurat.filt <- subset(human.seurat.filt, idents = "Spleen")
blood.seurat.filt <- subset(human.seurat.filt, idents = "Blood")

```


### Normalise datasets 
```{r normalise_datasets}

################################################
# Normalise, scale & find variable genes 
################################################


datasets.list <- list(human.seurat.filt, spleen.seurat.filt,  blood.seurat.filt)
names(datasets.list)  <- c("Complete", "Spleen", "Blood")

for(i in seq_along(datasets.list)){
  
  # Normalise RNA data by logNormalize
  datasets.list[[i]] <- NormalizeData(datasets.list[[i]],
                                      normalization.method = "LogNormalize",
                                      scale.factor = 10000,
                                      verbose = TRUE)
  
  
  # Find variable genes
  datasets.list[[i]] <- FindVariableFeatures(datasets.list[[i]],
                                             nfeatures = 1000,
                                             verbose = TRUE)
  
  # Scale data
  datasets.list[[i]] <- ScaleData(datasets.list[[i]],
                                  vars.to.regress = c("nCount_RNA", "percent.mito.recalc"),
                                  model.use = "linear",
                                  do.scale = TRUE,
                                  do.center = TRUE,
                                  verbose = TRUE)
  
  
}




```


## Unbiased Clustering {.tabset}

### Clustering
```{r Initial_clustering}

# Set seed for reproducibility
set.seed(42)

#################################################################
# Determine statistically significant principal components
#################################################################

human.seurat.filt <- datasets.list[[1]]

human.seurat.filt <- RunPCA(human.seurat.filt, 
                            features = human.seurat.filt@assays$RNA@var.features,
                            npcs = 50,
                            verbose = TRUE, 
                            rev.pca = TRUE,
                            seed.use = 42)

# Use quick approximation of significant PCs
ElbowPlot(human.seurat.filt, ndims = 30)
dev.copy(pdf, "output/QC/PC_elbowplot.pdf")
dev.off()


# NOTE: This process can take a long time for big datasets, comment out for
# expediency.  More approximate techniques such as those implemented in
# PCElbowPlot() can be used to reduce computation time

if(long.compute){
  
  human.seurat.filt <- JackStraw(human.seurat.filt,
                                 dims = 50)
  human.seurat.filt <- ScoreJackStraw(human.seurat.filt,
                                      dims = 1:50)
  
  JackStrawPlot(human.seurat.filt, dims = 1:30)
  
  dev.copy(pdf, "output/QC/JackStrawPlot.pdf")
  dev.off()
  
}

# PC 24 = last sig dim
# therefore use 20 dims



###################################
# PCA and cluster identification
###################################


for(i in seq_along(datasets.list)){
  
  # Create output dir
  output.path <- paste0("output/figures/PCA_", names(datasets.list)[i])
  
  
  if(!dir.exists(output.path)){
    dir.create(output.path, 
               recursive = TRUE)}
  
  # Run PCA
  datasets.list[[i]] <- RunPCA(datasets.list[[i]], 
                               features = datasets.list[[i]]@assays$RNA@var.features,
                               npcs = 20,
                               verbose = TRUE, 
                               rev.pca = TRUE,
                               seed.use = 42)
  
  # Plot PCA projection
  print(PCAPlot(datasets.list[[i]], 
          dims = c(1, 2),
          pt.size = 2,
          group.by = "organ"))
  
  dev.copy(pdf, paste0(output.path, "/PCA_by_organ_Dim1_2.pdf"))
  dev.off()
  
  print(PCAPlot(datasets.list[[i]], 
          dims = c(2, 3),
          pt.size = 2,
          group.by = "organ"))
  
  dev.copy(pdf, paste0(output.path, "/PCA_by_organ_Dim2_3.pdf"))
  dev.off()
  
  
  
  # Find neighbours
  datasets.list[[i]] <- FindNeighbors(datasets.list[[i]],
                                      dims = 1:20)
  
  # Find clusters
  datasets.list[[i]] <- FindClusters(datasets.list[[i]],
                                     resolution = 0.1)
  
}





```



### UMAP and tSNE dim Reduction
```{r Dim_reduction}



for(i in seq_along(datasets.list)){
  
  # Set assay 
  DefaultAssay(datasets.list[[i]]) <- "RNA"
  
  # Set Idents
  Idents(datasets.list[[i]]) <- datasets.list[[i]]@meta.data$organ
  
  
  
  
  ##########
  # TSNE
  ##########
  
  # Create output dir
  output.path <- paste0("output/figures/tSNE_", names(datasets.list)[i])
  
  
  if(!dir.exists(output.path)){
    dir.create(output.path, 
               recursive = TRUE)}
  
  
  
  
  datasets.list[[i]] <- RunTSNE(datasets.list[[i]],
                                reduction = "pca",
                                dims = 1:20,
                                perplexity = 30,
                                method = "FIt-SNE",
                                seed.use = 42,
                                dim.embed = 2)
  
  print(TSNEPlot(datasets.list[[i]],
           label = TRUE, 
           group.by = "organ"))
  
  dev.copy(pdf,  paste0(output.path, "/tSNE_by_organ.pdf"))
  dev.off()
  
  
  
  print(TSNEPlot(datasets.list[[i]],
           label = TRUE, 
           group.by = "seurat_clusters"))
  
  dev.copy(pdf, paste0(output.path, "/tSNE_by_seurat_clusters.pdf"))
  dev.off()
  
    print(TSNEPlot(datasets.list[[i]],
           label = TRUE, 
           group.by = "donor"))
  
  dev.copy(pdf,  paste0(output.path, "/tSNE_by_donor.pdf"))
  dev.off()
  
  ##########
  # UMAP
  ##########
  
  # Create output dir
  output.path <- paste0("output/figures/UMAP_", names(datasets.list)[i])
  
  
  if(!dir.exists(output.path)){
    dir.create(output.path, 
               recursive = TRUE)}
  
  
  datasets.list[[i]] <- RunUMAP(datasets.list[[i]],
                                reduction = "pca",
                                dims = 1:20,
                                umap.method = "uwot", 
                                n.neighbors = 45, # 5 to 50 // 45
                                min.dist = 0.5, # Sensible values are in the range 0.001 to 0.5 // 0.5
                                seed.use = 42)
  
  # Larger values = global structure
  # small value = local structure
  
  print(UMAPPlot(datasets.list[[i]],
           label = T,
           group.by = "organ"))
  
  dev.copy(pdf, paste0(output.path, "/UMAP_by_organ.pdf"))
  dev.off()
  
  print(UMAPPlot(datasets.list[[i]],
           label = TRUE, 
           group.by = "seurat_clusters"))
  
  dev.copy(pdf, paste0(output.path, "/UMAP_by_seurat_clusters.pdf"))
  dev.off()
  
  # Original Author clusters
  print(UMAPPlot(datasets.list[[i]],
           label = TRUE, 
           group.by = "donor"))
  
  dev.copy(pdf, paste0(output.path, "/UMAP_by_donor.pdf"))
  dev.off()
  
}









```


### Export seurat objects
```{r export_seurat_RDS}

for(i in seq_along(datasets.list)){

saveRDS(datasets.list[[i]], file = paste0("Exported_RDS_files/Human_", names(datasets.list)[i], "_seurat_filt.rds"))

  
}


```




## Imputation {.tabset}

### Imputation calculation
```{r seurat_imputation}


for(i in seq_along(datasets.list)){

# Impute gene expression 
datasets.list[[i]] <- RunALRA(datasets.list[[i]], 
                             genes.use = rownames(datasets.list[[i]]))

#The matrix went from 7.41% nonzero to 61.04% nonzero

# set assay 
DefaultAssay(datasets.list[[i]]) <- "RNA"

saveRDS(datasets.list[[i]], file = paste0("Exported_RDS_files/Human_", names(datasets.list)[i], "_seurat_filt_imputed.rds"))


}


```

